### 2012-07-20 Doc/VRIPHYS2012/vriphys2012.tex
modifs & comments
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@430 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\title[Short title]%
      {Bezier shell finite element for interactive medical simulation} % TODO

@@

This problem has been faced by the computer graphics community but for different applications.
For instance, many related work faces the problem of the simulation of cloth using mass-spring models \cite{Volino2009} or bending models \cite{Grinspun2003,Choi2007}. %% Todo verify.... 
However, these models are discrete so their behavior depends on the mesh and their parameters are not easily measurable.
This is completely acceptable for some computer graphics application, where the parameters can be tuned to obtain a realistic animation.
In the case of medical simulation, more physics-based modeling is needed.

Methods based on continuum mechanics and namely theory of elasticity gained on popularity in computer graphics and real-time simulators ever since Terzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.  %% Do we
While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling \footnote{TODO: do we need ref. here?  yes but for medical applications} of deformations, they are still frowned upon in the area of thin structures. 
Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
This theory can be integrated using Finite Elements, but there are two drawbacks: (i) the continuum equations requires second spatial derivations that can not be handled with linear interpolation and (ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).

The usual argument against non-linear FEM based methods is that they are computationally too expensive. 
This is indeed true if lots of elements are used to discretize the object.
A classical strategy, is to use high-polygonal mesh for rendering or collision detection and response, while maintaing a reasonably low number of elements for the mechanics.
The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynoms.
This type of interpolation is necessary because bending continuum model
(based on Kirchhoff-Love theory for thin plates)  requires the computation of second spatial derivatives.
In \cite{Comas2010c}, Comas and al.  use an hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
The element is piloted by 3 nodes on which 6 degrees of freedom are defined. 
From this point of view, our work is derived from this approach but we address several noted drawbacks: the formulation misses one rotational degree of freedom, the cubic interpolation is not symmetric and produces $C^0$  discontinuities between elements on edges.
Ubach and al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
So the stiffness forces computed on each element not only depends on nodes of the element but also on all the adjacent nodes. 
This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.

@@

\section{Element Geometry} % {{{

\subsection{B\'ezier Triangle} % {{{

@@

B\'ezier triangles were designed in sixties by French mathematician de
Casteljau. As with other B\'ezier or B-spline surfaces the B\'ezier triangle is
defined by a mesh of control points that do not necessarily lie on the
surface. The general $n$-th order B\'ezier triangle requires $(n + 1)(n +
2)/2$ control points and it's surface is defined as:

@@

\subsection{Control Points} % {{{

The B\'ezier triangles are defined by a mesh of control points, however the
simulated object is usually described by a triangular mesh with 6 degrees of freedom per node (position \& rotation).
We construct
the initial B\'ezier mesh by taking into account the curvature of the triangular
mesh. Note that special care has to be taken to maintain
the continuity across the nodes and edges. During the initialization phase
we partially employ the method described in \cite{Ubach2010} to maintain
$C^0$ continuity on the edges. Each of the control points on the edge is
computed as the intersection of:

@@

We do not repeat this procedure during the simulation. For fast update of
the control points we rigidly attach each edge point to nearest corner point (see Figure
\ref{fig-segments} for the correspondence between edge points and
associated corner points).

\Figure{0.8\linewidth}{segments}
{Correspondence between edge points and associated corner points shown in
blue.}

@@

In each step of the simulation we recompute the new position based on 
current orientation of the corner node:

@@

@@

$C^1$ continuity on the edges can also be maintained if special care is taken
when computing the position of the central point (see \cite{Ubach2010}). For
simplicity we compute the central point like this:

@@

@@

\noindent
thus computing the function of deflection $u$.

>>>>

%%% ADD COMMENTS=
\usepackage{color} 
\newcommand{\CD}[1]{{\color{magenta}{\textbf{C: #1}}}}

\title[Short title]%
      {Bezier shell finite element for interactive surgical simulation} % TODO

@@

This problem has been faced by the computer graphics community but for different applications.
For instance, many related work addresses the problem of the simulation of cloth using mass-spring models \cite{Volino2009} or bending models \cite{Grinspun2003,Choi2007}. %% Todo verify.... 
However, these models are discrete so their behavior depends on the mesh and their parameters are not easily measurable.
This is completely acceptable for some computer graphics application, where the parameters can be tuned manually to obtain a realistic animation.
In the case of medical simulation, more physics-based modeling is needed.

Methods based on continuum mechanics and namely theory of elasticity gained on popularity in computer graphics and real-time simulators ever since Terzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.  %% Do we
While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling \footnote{TODO: do we need ref. here?  yes but for medical applications} of deformations, they are still frowned upon in the area of thin structures. 
Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
This theory was integrated numerically \CD{Todo citation of a shell book} using Finite Elements, but there are two issues for obtaining interactive frame rates: 
(i) the continuum equations requires second spatial derivations that can not be handled with linear interpolation and 
(ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).

The usual argument against non-linear FEM based methods is that they are computationally too expensive. 
This is indeed true if lots of elements are used to discretize the object.
A classical strategy, is to use high-polygonal meshes for rendering or collision detection and response, while maintaing a reasonably low number of elements for the mechanics.
The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynoms.
This type of interpolation is necessary because bending continuum model
(based on Kirchhoff-Love theory for thin plates)  requires the computation of second spatial derivatives.
In \cite{Comas2010c}, Comas and al.  use an hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
The element is piloted by 3 nodes on which 6 degrees of freedom are defined. 
From this point of view, our work is derived from this approach but we address several noted drawbacks: the formulation misses one rotational degree of freedom, the cubic interpolation is not symmetric and produces $C^0$  discontinuities between elements on edges.
Ubach and al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
So the stiffness forces computed on each element not only depends on nodes of the element but also on all the adjacent nodes. 
This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.

@@

\section{Element kinematics} % {{{

\CD{intro: we propose an interpolation with a high degree of derivability while maintaining a reasonably low number of degrees of Freedom.
We introduce a new way of building the interpolation by using two stages: 
(i) a B\'ezier interpolation is defined with cubic polynomial functions 
(ii) a kinematic relation between nodes (6DOFs) defined at the vertices and the control points of the B\'ezier Triangle}

\subsection{B\'ezier Triangle Interpolation} % {{{

@@

B\'ezier triangles were designed in sixties by French mathematician de Casteljau.  % SURE ? What about Bezier himself ?
As with other B\'ezier or B-spline surfaces the B\'ezier triangle is
defined by a mesh of control points that do not necessarily lie on the
surface. The general $n$-th order B\'ezier triangle requires $(n + 1)(n +
2)/2$ control points and it's surface is defined as:

@@

\subsection{Kinematic between Nodes and Control Points } % {{{

The B\'ezier triangles are defined by a mesh of control points, however the
simulated object is usually described by a triangular mesh with 6 degrees of freedom per node (position \& rotation).
%
\CD{I would not say that it is "usual"... => I would say that 10 points x 3 Dofs  = 30 Dofs for a bezier triangle $\rightarrow$ it's quite a lot ! 
We propose to reduce the degrees of freedom by using 3 nodes with 6 dofs = 18 Dofs }
%
We constructthe initial B\'ezier mesh by taking into account the curvature of the triangular
mesh. Note that special care has to be taken to maintain
the continuity across the nodes and edges. During the initialization phase
we partially employ the method described in \cite{Ubach2010} to maintain
$C^0$ continuity on the edges. Each of the control points on the edge is
computed as the intersection of (\CD{Need to do a figure of this}):

@@

We do not repeat this procedure during the simulation. 
\CD{Why ?... do we argue that it is only for having fast update ?  } 
For fast update of
the control points we rigidly attach each edge point to nearest corner point (see Figure
\ref{fig-segments} for the correspondence between edge points and
associated corner points).

\Figure{0.8\linewidth}{segments}
{Correspondence between edge points and associated corner points shown in
blue. \CD{Not clear !!}}

@@

In each step of the simulation we recompute the new position based on 
current orientation of the corner node:
\CD{Other possible explanation: we apply the rigid transformation of frame $i$ on control point $j$ }

@@

\CD{To be consitent with what I write in following, we could introduce a notation for the dofs of the nodes:  $\mathbf{x}_i$ for the position and $\theta_i$ for the rotation of the node $i$
Then, the previous equation would be:
\begin{equation}
    P_j = \mathbf{x}_i + P_{i-j} = P_i + R(\theta_i) \overline{P_{i-j}}
\end{equation}
where $ R(\theta_i)$ is the rotation matrix for the angular position of node $i$...
}

@@

$C^1$ continuity on the edges can also be maintained if special care is taken
when computing the position of the central point (see \cite{Ubach2010}). 
For simplicity we compute the central point like this: 
\CD{Here it is not clear that with this formulation we do not maintain $C^1$ Continuity...}

@@

\CD{It's more about derivatives of the kinematic relation, right ?}

@@

\CD{I would use the following notation: small variation of the node $j$ position:   $\delta \mathbf{x}_j$ for the translation and $\delta \boldsymbol{\theta}_j$ for the rotation.
This small variation of the position creates a small variation of the control points $i$ position $\delta P_i$ that are on the edges:
$$
\delta P_i =  \delta \mathbf{x}_j +  P_{i-j} \times \delta \boldsymbol{\theta}_j
$$
 Variation of node $10$ will be:
$$
\delta P_{10} =  \sum_{j=1}^3  \frac{1}{3}(\delta \mathbf{x}_j +  P_{10-i} \times \delta \boldsymbol{\theta}_i)
$$
%
Thus, finally, it will creates a variation of the point position on the surface $\delta T$...
$$
\delta T =   \sum_{i=1}^{10} n_i  \delta P_i
$$
which makes:
\begin{equation}
    \begin{split}
\delta T =  (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 \\
 + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 \\
 + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 \\
+  \sum_{i=4}^9  \sum_{j=1}^3 n_i P_{i-j} \times \delta \theta_j \\
 +  n_{10} \frac{1}{3} \sum_{j=1}^3 P_{10-j} \times \theta_j
  \end{split}
\end{equation}
%
}

\noindent
thus computing the function of deflection $u$ 
\CD{ or $\delta T$ }.
